---
title: "Project: Exploring Seabird Data Over Time"
format: html
execute: 
  echo: false
  warning: false
  message: false
embed-resources: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)     # package for general data cleaning/tidyness
library(ggplot2)       # for creating my visualizations
library(paletteer)     # for color scales
library(reshape2)      # for creating the heatmap
library(ggthemes)      # for adding a new continuous color scale to the heatmap

```


## Data Source: 

> The data I collected is coming from a recent TidyTuesday repository. It is a collection of seabird data from 1969 to 1990, and it includes four different datasets. One for the observations of birds by species, one for the ship details/traveling conditions of the record ID, one for sea states, and one for beaufort scales. 

## Data Description:

> The birds dataset includes the record ID (numeric), bird name and scientific name as string variables, bird count (numeric), and behaviors (following ship, sitting on water, feeding) as boolean (TRUE/FALSE) values. The ships data set includes corresponding information on the record ID (numeric), date, time, latitude/longitude, activity, atmospheric/barometric measurements, season, and observer. The sea states data set includes the sea state class (corresponding to the ships data set) sea state verbal description, and wave meters min and waves meter max. The beaufort scale data set includes the wind speed class correpsonding to the ships data set (discrete), wind speed verbal description, and the wind speed knots min/max. With this data, I am seeking to answer the following mini-research questions. 

## Research Questions:

> What season are all seabirds most and least frequently observed in the Eastern hemisphere?

> What sea conditions are seabirds most frequently spotted (sea state)? 

## Data Cleaning

To clean the data, I first examined all four datasets, and found that only the "birds" and "ships" data would be meaningful for my analysis. I then selected only relevant variables to use from the birds data, such as the record id, observation id, and count (total number of birds observed per observation). I then found a matching variable in the birds and ships data sets and used it to join both together. This gave me additional metadata associated with each observation that pertained to my research questions. Next, I had to convert two variable (sea state and wind state) to factors, so I could treat them as distinct levels (sea state 1, 2, 3, etc.) instead keeping them on a continuous scale.


```{r}
#| label: data-cleaning

# read in ships data to join observation data to birds data
ships <- read_csv("data-raw/ships.csv")

# read in raw birds observation data
bird_count <- read_csv("data-raw/birds.csv")|>
  select(bird_observation_id,
         record_id,
         count)     # selecting only necessary variables

bird_count <- bird_count|>
  left_join(ships,
            by = "record_id")     # join the ship/excursion metadata from ships csv to matching record in bird data

bird_count <- bird_count|>      
  mutate(wind_speed_class = as.factor(wind_speed_class),  # change wind state scale to factor
         sea_state_class = as.factor(sea_state_class))    # change sea state scale to factor

write_csv(bird_count, "data-clean/bird_count.csv")

```

```{r}
#| label: read-data

# read the cleaned, joined data from the data-clean folder to be used in visualization creation
birds <- read_csv("data-clean/bird_count.csv")


```

## Visualization 1 - Seasons Barplot

This bar plot investigates if there are any seasons where seabirds are observed the most across the entire time series of observations from 1969-1990. The x-axis is mapped to season (winter, spring, summer, fall), and the y-axis is the total count of all birds. From here we can clearly see that seabirds are most frequently observed in the spring, followed closely by summer. This makes sense as rising sea and coastal temperatures in spring, along with increased sunlight availability, can lead to more abundance of phytoplankton, causing a trophic cascade of food/fish availability in the water column for seabirds. Migratory seabirds follow open-ocean hunting patterns year-round, but it may be easier to observe these behaviors in the spring and summer when such resources are abundant for fish, as compared to the winter and autumn seasons.  



```{r}
#| label: viz-1
#| fig-cap: "Figure 1: Bar plot comparing the total counts for seabirds across all species by calendar season. Data: TidyTuesday (Github)"
#| fig-alt: "A bar plot showing the total amount of seabirds observed across the four seasons from 1969 to 1990. Spring has the highest overall value for observations, with roughly 800,000 total observations, followed by summer with roughly 700,000 total observations. The bars for autumn and winter are close to each other in height, each close to 175,000 total observations. This suggests that summertime has the highest frequency for seabird sightings, whereas seabirds are less likely to observed in autumn and winter."
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

birds|>
  drop_na(count, season)|>         # drop na values for count and season
  ggplot(mapping = aes(x = factor(season), y = count, fill = season)) + 
  geom_col() +  
  labs(x = "Season",
       y = "Total Seabird Obsevrations",
       title = "Seabird Observations Across Seasons (1969-1990)") +
  scale_x_discrete(labels = c("autumn" = "Autumn",
                              "spring" = "Spring",
                              "summer" = "Summer",
                              "winter" = "Winter")) +      # relabel seasons
  scale_fill_paletteer_d("calecopal::lupinus") +
  theme_bw() +        # set theme
  theme(legend.position = "none")        # remove legend from plot
  

```

## Visualization 2 - Sea Conditions Heatmap

This heatmap investigates if there is are any patterns in overall sea state and the frequency of seabird observations. The x-axis, sea state class, is a measure of sea conditions (waves, swell, etc.), and higher values indicate rougher waters. Similarly, the y-axis, wind state class, is a measure of wind force, with higher values indicating more blustery seas. Both use the Beaufort weather scale. Color is mapped to the log-transformed total count of seabirds across all observations from 1969-1990, with reddish orange representing more observations and purplish blue indicating less observations. Compared to neighboring tiles, there is a distinct pinkish-orange clustering of tiles in the center of distribution, around sea state class 3-4 and wind state class 6-7. This could indicate that moderate weather conditions yield a higher frequency of overall seabird observations.  



```{r}
#| label: viz-2
#| fig-cap: "Figure 2: Heatmap displaying the intensity of seabird observations by sea state class and wind state class. Data: TidyTuesday (Github)"
#| fig-alt: "A heatmap displaying the relationship between sea conditions and the log-transformed seabird observation intensity. Sea state class, a measure of the general condition of the ocean's surface (Beuafort scale 1-6), is mapped to the x-axis, and wind state class, a measure of wind speed and strength (Beaufort scale 0-12), is mapped to the y-axis. Log-transformed observation intensity is mapped to a continuous color gradient. We see a small clustering of color centered around sea state class 3-4 and wind state class 6-7. This indicates that moderate sea and wind conditions are attributed to a higher frequency of overall seabird observations."
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

birds|>
  drop_na(count,
          wind_speed_class,
          sea_state_class)|>          # drop na values from count, wind speed, and sea state
  ggplot(mapping = aes(x = sea_state_class,
                       y = wind_speed_class,
                       fill = log(count))) +      # log-transform the bird count so it is easier to visualize
  geom_tile() +
  scale_fill_paletteer_c("ggthemes::Sunset-Sunrise Diverging") + # add sunset color scale
  labs(title = "Seabird Observation Intensity for Sea State and Wind Classes",
       x = "Sea State Class",
       y = "Wind Speed Class") +            # add appropriate labeling
  theme_bw()         # same theme from previous plot

```
